### 2015-03-14 background.tex
updates to background

\subsection{Concurrency in the Linux Kernel}
\label{concurrency}

@@

Figure~\ref{data_race_example} is an example of a data race between two entry points of a Linux module. The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}. Entry point \texttt{ep1} writes the value 5 to \texttt{dev->resource}, and entry point \texttt{ep2} writes the value 3 to \texttt{dev->resource} if the value of \texttt{dev->resource} is not 5. A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behaviour.

@@

static void ep2 (struct shared *dev) {
  if (dev->resource != 5)
    dev->resource = 3;
}
\end{lstlisting}
\caption{Simple data race in a Linux module}
\label{data_race_example}
\end{figure}

@@

static void ep2 (struct shared *dev) {
lock(&dev->lock);
  if (dev->resource != 5)
    dev->resource = 3;
unlock(&dev->lock);
}
\end{lstlisting}
\end{minipage}
\caption{Using locks to eliminate the data race}
\label{lock_example}
\end{figure}

The most common technique for eliminating data races is using a \emph{locking} mechanism. Locking a set of program statements that access a shared resource creates a \emph{critical section}, i.e. source code that cannot be executed by more than one thread simultaneously. Figure~\ref{lock_example} shows how to use a lock to eliminate the data race in the previous example.

Lock-based synchronisation is a powerful tool for preventing data races in drivers, but careless use of locks has multiple pitfalls~\cite{sutter2005software}. Coarse-grained locking can severely hurt performance, because it limits the opportunity for concurrency. On the other hand, fine-grained locking can easily lead to \emph{deadlocks}, a system-halting situation in which no active thread can progress, because it is waiting to acquire a lock that is held by another blocked thread. Although the Linux kernel provides sophisticated lock-free synchronisation mechanisms~\cite{corbet2005linux}, in this work we choose to focus on reasoning about locks.

\subsection{Limitations in Reasoning about Thread Interleavings}

A fundamental concept in concurrency analysis is Lamport's \emph{happens-before} relation~\cite{lamport1978time}, which denotes a partial-order between all events during a concurrent execution of a program. The happens-before relation can be defined as follows:

\begin{definition}
\label{definition:datarace}
Let $E_1$ and $E_2$ be two events during a program's concurrent execution. If $E_1$ and $E_2$ are in the same thread and $E_1$ occurs before $E_2$, or if $E_1$ and $E_2$ are in different threads and there is a synchronisation mechanism that prevents a thread schedule in which $E_2$ occurs before $E_1$, then $E_1 \rightarrow E_2$, where the symbol $\rightarrow$ denotes the happens-before relation.
\end{definition}

Happens-before is transitive, i.e. for any three events $E_1$, $E_2$ and $E_3$, if $E_1 \rightarrow E_2$ and $E_2 \rightarrow E_3$, then $E_1 \rightarrow E_3$. If two events $E_1$ and $E_2$ occur in different threads, and neither $E_1 \rightarrow E_2$, nor $E_2 \rightarrow E_1$ holds, then $E_1$ and $E_2$ happen concurrently, which indicates a data race if the two events access the same shared resource and at least one of the accesses is a modifier. Figure~\ref{happens_before} shows an example of applying happens-before reasoning in two concurrently running entry points of a driver. Each of the three statements \{A, B, C\} in thread 1 is executed before the next one, and thus all of them are ordered by the happens-before relation. The same applies for each of the four statements \{D, E, F, G\} in thread 2. Statement C in thread 1 releases the lock, and subsequently allows D in thread 2 to acquire the same lock. Thus, $C \rightarrow D$. This means that for this specific thread interleaving the set of statements \{A, B, C\} happens before the set of statements \{D, E, F, G\}.

\begin{figure}[htbp]
\centering
\includegraphics[width=.95\linewidth]{img/happens_before.pdf}
\caption{Example showcasing the main limitation of happens-before.}
\label{happens_before}
\end{figure}

The above example showcases the main limitation of happens-before. If the technique explores \emph{only} this specific thread interleaving, it will miss a potential data race between statements B and G, because $B \rightarrow G$ in this specific schedule. Thus, a tool based on happens-before requires a scheduler that produces as many different thread interleavings as possible to increase execution path coverage~\cite{savage1997eraser}. Achieving full coverage with tools based on happens-before is infeasible though, because of the exponential number of possible thread interleavings in realistic concurrent programs. Furthermore, attempting to explore a large number of thread schedules quickly leads to scalability issues~\cite{musuvathi2008finding}. Thus, previous work based on happens-before typically explore paths under a certain bound, such as the number of allowed context switches~\cite{qadeer2004kiss}.

>>>>

\subsection{Concurrency in the Linux Kernel}
\label{bg:concurrency}

@@

Figure~\ref{fig:data_race_example} is an example of a data race between two entry points of a Linux module. The resource \texttt{dev->resource} is shared between the two entry points \texttt{ep1} and \texttt{ep2}. Entry point \texttt{ep1} writes the value 5 to \texttt{dev->resource}, and entry point \texttt{ep2} writes the value 3 to \texttt{dev->resource} if the value of \texttt{dev->resource} is not 5. A data race can occur if \texttt{ep1} and \texttt{ep2} run concurrently: the read and write accesses to the shared resource may execute in arbitrary order leading to nondeterministic behaviour.

@@

static void ep2 (struct shared *dev) {
  if (dev->resource != 5)
    dev->resource = 3;
}
\end{lstlisting}
\caption{Simple data race in a Linux module}
\label{fig:data_race_example}
\end{figure}

@@

static void ep2 (struct shared *dev) {
  lock(&dev->lock);
  if (dev->resource != 5)
    dev->resource = 3;
  unlock(&dev->lock);
}
\end{lstlisting}
\end{minipage}
\caption{Using locks to eliminate the data race}
\label{fig:lock_example}
\end{figure}

The most common technique for eliminating data races is using a \emph{locking} mechanism. Locking a set of program statements that access a shared resource creates a \emph{critical section}, i.e. source code that cannot be executed by more than one thread simultaneously. Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the data race in the previous example.

Lock-based synchronisation is a powerful tool for preventing data races in drivers, but careless use of locks has multiple pitfalls~\cite{sutter2005software}. Coarse-grained locking can severely hurt performance, because it limits the opportunity for concurrency. On the other hand, fine-grained locking can easily lead to \emph{deadlocks}, a system-halting situation in which no active thread can progress, because it is waiting to acquire a lock that is held by another blocked thread. Although the Linux kernel provides sophisticated lock-free synchronisation mechanisms~\cite{corbet2005linux}, in this work we focus on locks.

\subsection{Reasoning about Thread Interleavings}
\label{bg:happensbefore}

A fundamental concept in concurrency analysis is Lamport's \emph{happens-before} relation~\cite{lamport1978time}, which denotes a partial-order between all events during the execution of a concurrent program. Happens-before can be defined as follows:

\begin{definition}
\label{definition:datarace}
Let $E_1$ and $E_2$ be two events during a program's concurrent execution. If $E_1$ and $E_2$ are in the same thread and $E_1$ occurs before $E_2$, or if $E_1$ and $E_2$ are in different threads and there is a synchronisation mechanism (e.g. locking) that prevents a thread schedule in which $E_2$ occurs before $E_1$, then $E_1 \rightarrow E_2$, where the symbol $\rightarrow$ denotes the happens-before relation.
\end{definition}

Happens-before is transitive, i.e. for any three events $E_1$, $E_2$ and $E_3$, if $E_1 \rightarrow E_2$ and $E_2 \rightarrow E_3$, then $E_1 \rightarrow E_3$. If two events $E_1$ and $E_2$ occur in different threads, and neither $E_1 \rightarrow E_2$, nor $E_2 \rightarrow E_1$ holds, then $E_1$ and $E_2$ happen concurrently, which indicates a data race if the two events access the same shared resource and at least one of the accesses is a modifier.

\begin{figure}[htbp]
\centering
\includegraphics[width=1\linewidth]{img/happens_before.pdf}
\caption{Example showcasing the main limitation of happens-before}
\label{fig:happens_before}
\end{figure}

Figure~\ref{fig:happens_before} shows an example of applying happens-before reasoning in two concurrently running entry points. Each of the three statements \{A, B, C\} in thread 1 is executed before the next one, and thus all of them are ordered by the happens-before relation. The same applies for each of the four statements \{D, E, F, G\} in thread 2. Statement C in thread 1 releases the lock, and subsequently allows D in thread 2 to acquire the same lock. Thus, $C \rightarrow D$. This means that for this specific thread interleaving the set of statements \{A, B, C\} happens before the set of statements \{D, E, F, G\}.

The above example also showcases the main limitation of happens-before. If the technique explores \emph{only} this specific thread interleaving, it will miss a potential data race between statements B and G, because $B \rightarrow G$ in this specific schedule. Thus, a tool based on happens-before requires a scheduler that produces as many different thread interleavings as possible to increase execution path coverage~\cite{savage1997eraser}. Achieving full coverage with tools based on happens-before is infeasible though, because of the exponential number of possible thread interleavings in realistic concurrent programs. Furthermore, attempting to explore a large number of thread schedules quickly leads to scalability issues~\cite{musuvathi2008finding}, a problem known as state space explosion. Thus, previous work based on happens-before typically explore paths under a certain bound, such as the number of allowed context switches~\cite{qadeer2004kiss}.

\subsection{Reasoning about Locks}
\label{bg:lockset}

An alternative approach to happens-before, known as \emph{lockset analysis}, resolves around reasoning about the locking behaviour of concurrent programs. Lockset analysis has its roots in Eraser~\cite{savage1997eraser}, a dynamic data race detector that tracks the set of locks that \emph{consistently} protect a memory location during program execution. If that lockset ever becomes empty, the analysis reports a potential data race on that memory location. This is because an inconsistent lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.

\begin{figure}[htbp]
\centering
\includegraphics[width=1\linewidth]{img/lockset.pdf}
\caption{Applying lockset analysis on two threads}
\label{fig:locksets}
\end{figure}

Figure~\ref{fig:locksets} shows an example of applying lockset analysis on two concurrently running threads.

Lockset analysis avoids reasoning about arbitrary thread interleavings, and thus has the potential to scale well in realistic concurrent programs. The technique, though, suffers from imprecision (i.e. can report many false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, as a classic dynamic analyser, Eraser's bug finding ability is limited to the execution paths that the tool explores.

### 2015-03-14 intro.tex
updates to background

We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction and significantly accelerate \corral~\cite{lal2012solver}, a state-of-the-art bug finder that is used by Microsoft to analyse Windows device drivers. We have prototyped this approach in \whoop, a practical tool for automatic concurrency verification, which we used to analyse 10 modules available in the Linux distribution.

>>>>

We then show that by exploiting the race-freedom guarantees that our static analysis provides, we can achieve a sound partial-order reduction and significantly accelerate \corral~\cite{lal2012solver}, a state-of-the-art bug finder that is used by Microsoft to analyse device drivers. We have prototyped this approach in \whoop, a practical tool for automatic concurrency verification, which we used to analyse 10 modules available in the Linux distribution.

### 2015-03-14 theory.tex
updates to background

Our technique resolves around reasoning about the \emph{lock sets} of a device driver. Lock set analysis has its roots in Eraser~\cite{savage1997eraser}, a dynamic data race detector that tracks the set of locks that consistently protect a memory location during program execution. If that lock set ever becomes empty, the tool reports a potential data race. This is because an inconsistent lock set suggests that a memory location can be accessed simultaneously by two or more threads. Lock set analysis avoids reasoning about arbitrary thread interleavings, and thus can scale well in realistic concurrent programs, but suffers from imprecision (i.e. can report many false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, as a classic dynamic analyser, Eraser's bug finding ability is limited to the execution paths that the tool explores. To counter these limitations, we apply the idea of Eraser's lock set analysis in a static verification context.

We propose \emph{static pair-wise lock set analysis}, a novel verification technique for proving data race freedom in device drivers. The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent driver, (ii) instrumenting it for lock set analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.

The proof sketch behind our static lock set analysis is given in Section~\ref{proof}. The verification technique is described in more detail in Section~\ref{method}. The prototype implementation of our approach as the practical tool \textsc{Whoop} is presented in Section~\ref{implementation}.

\subsection{Static lock set analysis proof sketch}
\label{proof}

We now formalise our approach to verifying data race freedom in a device driver by statically analysing its lock sets. Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver. Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all device drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one). Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking.

For each entry point $\mathit{ep}_{i}$, we compute its lock set $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$. Furthermore, for each entry point $\mathit{ep}_{i}$, we compute $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which maps each $m$ to true if and only if $\mathit{ep}_{i}$ writes to $m$ during some execution.

@@

A sketch of how the above theorem can be proved is as follows. Suppose there are in fact entry points $\mathit{ep}_{i}$ and $\mathit{ep}_{j}$ that can race on a memory location $m$. By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$. By the definition of a lock set, this means that $\ell$ is held during the access to $m$ by both entry points $ep_1$ and $ep_2$. As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

@@

Next, we split each pair into two distinct analysis regions: the \emph{logger}, which contains all the statements from the first entry point; and the \emph{checker}, which contains all the statements from the second entry point. The logger and the checker are then instrumented according to Table~\ref{tab:instrumentation}. The two regions directly contribute to the static lock set analysis and race checking as follows: the logger is responsible for recording all lock sets $\mathit{LS}$ maintained by its entry point; while the checker asserts for potential conflicts between its own entry point and a given lock set $\mathit{LS}$ from the logger. This whole process is summarised below:

\begin{description}
  \item[Logger] Initially, for each memory location $m$ accessed by the logger's entry point, the logger's lock set $\mathit{LS}\lbrack m\rbrack$ contains all locks $\ell$ in the device driver, while the write access set $W\lbrack m\rbrack$ is set to false. Also, the logger's current lock set $\mathit{LS}_{CR}$, which is the set of all locks \emph{currently} held by the logger's entry point, is initially empty. Each time the logger performs a read or write access to $m$, it updates its lock set $\mathit{LS}\lbrack m\rbrack$, with the intersection of $\mathit{LS}\lbrack m\rbrack$ and $\mathit{LS}_{CR}$. This process is known as \emph{lockset refinement}~\cite{savage1997eraser} and results into a $\mathit{LS}\lbrack m\rbrack$ that contains only the locks $\ell$ that consistently protect $m$ during the execution of the logger's entry point.

  \item[Checker] Initially, for each memory location $m$ accessed by the checker's entry point, the checker's lock set $\mathit{LS}\lbrack m\rbrack$ and write access set $W\lbrack m\rbrack$ are equal to the corresponding sets supplied at the end of the logger. Also, the logger's current lock set $\mathit{LS}_{CR}$ is initially empty. Each time the checker performs a write access to $m$, it asserts that the intersection between $\mathit{LS}_{CR}$ and $\mathit{LS}\lbrack m\rbrack$ is non-empty. If the assertion fails, it implies that the entry point of the logger and the entry point of the checker potentially race on $m$. In this case a counterexample is generated and reported to the user. To check a read access, the checker performs the aforementioned assertion only if the write access set $W\lbrack m\rbrack$ is true, thus avoids to report a read-read data race, which is inherently benign.
\end{description}

>>>>

To counter these limitations, we apply the idea of Eraser's lockset analysis in a static context.

We propose \emph{static pair-wise lockset analysis}, a novel verification technique for proving data race freedom in device drivers. The key idea behind our approach is that we can verify absence of data races by (i) deriving a sound \emph{sequential} model that \emph{over-approximates} the originally concurrent driver, (ii) instrumenting it for lockset analysis and race checking, and (iii) asserting that all write/read accesses to the same shared resource are consistently protected by at least one common lock. The immediate benefit is that our approach not only avoids reasoning about thread interleavings, and thus has the potential to scale well, but also allows the reuse of existing successful sequential verification techniques.

The proof sketch behind our static lockset analysis is given in Section~\ref{proof}. The verification technique is described in more detail in Section~\ref{method}. The prototype implementation of our approach as the practical tool \textsc{Whoop} is presented in Section~\ref{implementation}.

\subsection{Static lockset analysis proof sketch}
\label{proof}

We now formalise our approach to verifying data race freedom in a device driver by statically analysing its locksets. Let $\{\mathit{ep}_{1}, \mathit{ep}_{2}, \dotsc, \mathit{ep}_{n}\}$ be all the entry points of a driver, and let $\{\ell_{1}, \ell_{2}, \dotsc, \ell_{k}\}$ be all the locks used by this driver. Although we assume that the number of locks is finite and with a known bound, we argue that this assumption is realistic: all device drivers that we have studied so far in the Linux kernel have only a small number of locks (typically one). Indeed, the Linux device driver book~\cite{corbet2005linux} advocates the use of as few locks as possible to avoid introducing unnecessary complexity and synchronisation bugs that arise from careless fine-grained locking.

For each entry point $\mathit{ep}_{i}$, we compute its lockset $\mathit{LS}_{i}$ that associates each memory location $m$ to the set of locks that are always held when $m$ is accessed during the execution of $\mathit{ep}_{i}$. Furthermore, for each entry point $\mathit{ep}_{i}$, we compute $W_{i}\lbrack m\rbrack \rightarrow \{true, false\}$, which maps each $m$ to true if and only if $\mathit{ep}_{i}$ writes to $m$ during some execution.

@@

A sketch of how the above theorem can be proved is as follows. Suppose there are in fact entry points $\mathit{ep}_{i}$ and $\mathit{ep}_{j}$ that can race on a memory location $m$. By our hypothesis, there exists at least one lock, say $\ell$, which belongs to both $\mathit{LS}_{i}$ and $\mathit{LS}_{j}$. By the definition of a lockset, this means that $\ell$ is held during the access to $m$ by both entry points $ep_1$ and $ep_2$. As a result, $m$ \emph{must} be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

@@

Next, we split each pair into two distinct analysis regions: the \emph{logger}, which contains all the statements from the first entry point; and the \emph{checker}, which contains all the statements from the second entry point. The logger and the checker are then instrumented according to Table~\ref{tab:instrumentation}. The two regions directly contribute to the static lockset analysis and race checking as follows: the logger is responsible for recording all locksets $\mathit{LS}$ maintained by its entry point; while the checker asserts for potential conflicts between its own entry point and a given lockset $\mathit{LS}$ from the logger. This whole process is summarised below:

\begin{description}
  \item[Logger] Initially, for each memory location $m$ accessed by the logger's entry point, the logger's lockset $\mathit{LS}\lbrack m\rbrack$ contains all locks $\ell$ in the device driver, while the write access set $W\lbrack m\rbrack$ is set to false. Also, the logger's current lockset $\mathit{LS}_{CR}$, which is the set of all locks \emph{currently} held by the logger's entry point, is initially empty. Each time the logger performs a read or write access to $m$, it updates its lockset $\mathit{LS}\lbrack m\rbrack$, with the intersection of $\mathit{LS}\lbrack m\rbrack$ and $\mathit{LS}_{CR}$. This process is known as \emph{lockset refinement}~\cite{savage1997eraser} and results into a $\mathit{LS}\lbrack m\rbrack$ that contains only the locks $\ell$ that consistently protect $m$ during the execution of the logger's entry point.

  \item[Checker] Initially, for each memory location $m$ accessed by the checker's entry point, the checker's lockset $\mathit{LS}\lbrack m\rbrack$ and write access set $W\lbrack m\rbrack$ are equal to the corresponding sets supplied at the end of the logger. Also, the logger's current lockset $\mathit{LS}_{CR}$ is initially empty. Each time the checker performs a write access to $m$, it asserts that the intersection between $\mathit{LS}_{CR}$ and $\mathit{LS}\lbrack m\rbrack$ is non-empty. If the assertion fails, it implies that the entry point of the logger and the entry point of the checker potentially race on $m$. In this case a counterexample is generated and reported to the user. To check a read access, the checker performs the aforementioned assertion only if the write access set $W\lbrack m\rbrack$ is true, thus avoids to report a read-read data race, which is inherently benign.
\end{description}
